Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-21 Paper/manuscript_v2/manuscript_07_01.tex
Minor edits to sec 5.3

As the WBC predictor values were strictly positive and highly skewed, logarithm transformation is used for it. Continuous predictors were normalized to have zero mean and unit standard deviation. %Survival time was normalized to have zero mean for the logarithm of time. 
We assume a log-normal observation model for the observed survival time, $t_i$, with a function of the predictors, $f(\bm{x}_i):{\rm I\!R}^4 \to {\rm I\!R}$, as the location parameter, and $\sigma$ as the Gaussian noise: 
%
\begin{equation*}
  p(t_i)= \LogNormal(t_i \mid f(\bm{x}_i),\sigma^2).
\end{equation*}

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$ depending on predictors $x$ and hyperparameters $\theta=(\alpha,\ell)$,
%
\begin{equation*} 
f(\bm{x}) \sim \GP(c + \bm{\beta}\bm{x}, k(\bm{x},\bm{x}', \theta)),
\end{equation*}
where $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model.
% Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ are multivariate Gaussian distributed with mean function $\mu(\cdot)$ and covariance matrix $K$, where $\mu(\bm{x}_i)=c + \bm{\beta}\bm{x}_i$ and $K_{rs}=k(\bm{x}_r,\bm{x}_s,\theta)$, with $r,s=1,\dots,n$.
The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. A scalar lengthscale is used in the multivariate covariance function assuming an isotropic function.

@@

Figure \ref{ch5_fig21_posteriors_leukemia} shows estimated conditional functions of each predictor with all others fixed to their mean values. These posterior estimates correspond to the HSGP model with $m=10$ basis functions and $c=3$ boundary factor. There are clear  non-linear patterns and the right bottom subplot also shows that the conditional function associated with WBC has an interaction with TDI.
Figure~\ref{ch5_fig22_elpd_leukemia} shows the expected log predictive density \citep[ELPD; ][]{vehtari_2012,Vehtari+Gelman+Gabry:2017_practical} and time of computation as function of the number of univariate basis functions $m$ ($m^{\ast}=m^D$ in equation (\ref{eq_approxf_multi})) and boundary factor $c$. As the functions are smooth, a few number of basis functions and a large boundary factor are required to obtain a good approximation (Figure \ref{ch5_fig22_elpd_leukemia}-left); Small boundary factors are not allowed when large lengthscales, as can be seen in Figure \ref{fig6_lscale_vs_J_vs_c}. Increasing the boundary factor also significantly increases the time of computation (Figure \ref{ch5_fig22_elpd_leukemia}-right). With a moderate number of univariate basis functions ($m=15$), the HSGP model becomes slower than the exact GP model, in this specific application with $3$ input variables, as the total number of multivariate basis functions becomes $15^3 = 3375$ and is therefore quite high. 
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig22_elpd_leukemia.pdf}} 
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig23_time_leukemia.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 45mm 120mm 0mm, clip]{ch5_fig24_legend_leukemia.pdf}}
\caption{Expected log predictive density (ELPD; left) and time of computation in seconds per iteration (iteration of the HMC sampling method; right) as a function of the number of basis functions $m$ and boundary factor $c$.}
  \label{ch5_fig22_elpd_leukemia}
\end{figure}

>>>>

As the WBC predictor values were strictly positive and highly skewed, a logarithm transformation is used. Continuous predictors were normalized to have zero mean and unit standard deviation. %Survival time was normalized to have zero mean for the logarithm of time. 
We assume a log-normal observation model for the observed survival time, $t_i$, with a function of the predictors, $f(\bm{x}_i):{\rm I\!R}^4 \to {\rm I\!R}$, as the location parameter, and $\sigma$ as the Gaussian noise: 
%
\begin{equation*}
  p(t_i|f_i)= \LogNormal(t_i \mid f(\bm{x}_i),\sigma^2).
\end{equation*}

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i|f)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$ depending on predictors $x$ and hyperparameters $\theta=(\alpha,\ell)$,
%
\begin{equation*} 
f(\bm{x}) \sim \GP(c + \bm{\beta}\bm{x}, k(\bm{x},\bm{x}', \theta)),
\end{equation*}
where $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model.
% Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ are multivariate Gaussian distributed with mean function $\mu(\cdot)$ and covariance matrix $K$, where $\mu(\bm{x}_i)=c + \bm{\beta}\bm{x}_i$ and $K_{rs}=k(\bm{x}_r,\bm{x}_s,\theta)$, with $r,s=1,\dots,n$.
The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. A scalar lengthscale is used in the multivariate covariance function assuming an isotropic function.

@@

Figure \ref{ch5_fig21_posteriors_leukemia} shows estimated conditional functions of each predictor with all others fixed to their mean values. These posterior estimates correspond to the HSGP model with $m=10$ basis functions and $c=3$ boundary factor. There are clear  non-linear patterns and the right bottom subplot also shows that the conditional function associated with WBC has an interaction with TDI.
Figure~\ref{ch5_fig22_elpd_leukemia} shows the expected log predictive density \citep[ELPD; ][]{vehtari_2012,Vehtari+Gelman+Gabry:2017_practical} and time of computation as function of the number of univariate basis functions $m$ ($m^{\ast}=m^D$ in equation (\ref{eq_approxf_multi})) and boundary factor $c$. As the functions are smooth, a few number of basis functions and a large boundary factor are required to obtain a good approximation (Figure \ref{ch5_fig22_elpd_leukemia}-left); Small boundary factors are not appropriate for models for large lengthscales, as can be seen in Figure \ref{fig6_lscale_vs_J_vs_c}. Increasing the boundary factor also significantly increases the time of computation (Figure \ref{ch5_fig22_elpd_leukemia}-right). With a moderate number of univariate basis functions ($m=15$), the HSGP model becomes slower than the exact GP model, in this specific application with $3$ input variables, as the total number of multivariate basis functions becomes $15^3 = 3375$ and is therefore quite high. 
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig22_elpd_leukemia.pdf}} 
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig23_time_leukemia.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 45mm 120mm 0mm, clip]{ch5_fig24_legend_leukemia.pdf}}
\caption{Expected log predictive density (ELPD; left) and time of computation in seconds per iteration (iteration of the HMC sampling method; right) as a function of the number of basis functions $m$ and boundary factor $c$.}
  \label{ch5_fig22_elpd_leukemia}
\end{figure}
